
---
title: 'Retail Sales Analysis: Superstore Dataset'
author: "G. Turgen"
output:
  pdf_document: default
  html_document: default
---

# Introduction

This case study explores sales data from the Superstore dataset to uncover patterns in product performance, customer segments, and profitability trends. The goal is to derive actionable insights to improve revenue and optimize business strategies.

# Business Task

Analyze Superstore sales data to identify high-performing products, profitable customer segments, and regional sales trends. Use these insights to inform marketing, inventory, and pricing strategies.

# Stakeholders

- Regional Sales Directors
- Product Management Team
- Marketing & Strategy Department

# About the Data

The dataset includes fictional sales transactions from a U.S.-based retail store, containing details such as order date, product category, region, customer segment, sales, and profit. This dataset is commonly used for data visualization and business analytics training.

# ROCCC Assessment

- **Relevant**: Yes — closely aligned with the business questions.  
- **Original**: No — publicly available, not proprietary.  
- **Comprehensive**: Moderate — includes key variables (e.g., date, sales, region, segment).
- **Current**: Undated — no indication of real-world time frame.  
- **Cited**: Yes — sourced from Tableau Public / Kaggle.

# Data Cleaning

```{r packages}
library(tidyverse)
library(janitor)
library(lubridate)
library(skimr)
library(ggplot2)
library(readr)

```

## Load the data
```{r load the data}
superstore_df <- read_csv("/Users/dimitrid./Superstore Case Study/Sample - Superstore.csv") %>% 
  clean_names()
```

## Convert dates
```{r convert dates}
superstore_df <- superstore_df %>%
  mutate(order_date = mdy(order_date),
         ship_date = mdy(ship_date))
```


# Exploratory Analysis

## Product Performance

```{r top-selling products}
superstore_df %>%
  group_by(product_name) %>%
  summarize(total_sales = sum(sales, na.rm = TRUE)) %>%
  arrange(desc(total_sales)) %>%
  head(10)
```

* The highest-grossing products are large-ticket items, such as copiers and videoconferencing units.
* The top product alone (Canon Copier) accounts for over $60,000 in sales.
* Sales are heavily concentrated in a few high-value products. This suggests that while the overall product range is broad, a small subset drives the majority of revenue.

## Sales Trends Over Time Visualization

```{r monthly sales trend plot}
superstore_df %>%
  mutate(month = floor_date(order_date, "month")) %>%
  group_by(month) %>%
  summarize(monthly_sales = sum(sales, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = monthly_sales)) +
  geom_line(color = "steelblue") +
  labs(title = "Monthly Sales Trend", x = "Month", y = "Sales")
```

## Regional Performance Visualization

```{r sales by region plot}
superstore_df %>%
  group_by(region) %>%
  summarize(total_sales = sum(sales, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(region, -total_sales), y = total_sales, fill = region)) +
  geom_col() +
  labs(title = "Sales by Region", x = "Region", y = "Total Sales") +
  theme(legend.position = "none")
```

## Customer Segment Analysis

**Sales and Profit by Customer Segment:**

```{r customer segment performance}
superstore_df %>%
  group_by(segment) %>%
  summarize(
    total_sales = sum(sales, na.rm = TRUE),
    total_profit = sum(profit, na.rm = TRUE),
    avg_profit_margin = mean(profit / sales, na.rm = TRUE),
    order_count = n()
  ) %>%
  arrange(desc(total_sales))

```

While Consumers generate the most revenue, Home Office customers appear to be more profitable on average. This indicates that different customer types bring different types of value—volume vs. profitability—and may warrant distinct engagement strategies.

**Visualize Sales and Profit:**
```{r sales and profit plot}
superstore_df %>%
  group_by(segment) %>%
  summarize(
    total_sales = sum(sales, na.rm = TRUE),
    total_profit = sum(profit, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(total_sales, total_profit), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = segment, y = value, fill = metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Sales & Profit by Customer Segment", x = "Segment", y = "USD") +
  theme_minimal()

```

**Visualize Sales by Segment + Region:**

```{r sales by segment and region plot}
superstore_df %>%
  group_by(segment, region) %>%
  summarize(total_sales = sum(sales, na.rm = TRUE)) %>%
  ggplot(aes(x = region, y = segment, fill = total_sales)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Sales by Segment and Region", x = "Region", y = "Segment") +
  theme_minimal()

```

## Profitability Breakdown
This analysis will help you understand where profit is actually coming from, regardless of sales volume.
My goal is to identify which categories, sub-categories, or regions are driving or hurting profitability — even if their sales look good.

**Profitability by Category and Sub-Category:**
```{r category and sub-category profitability}
superstore_df %>%
  group_by(category, sub_category) %>%
  summarize(
    total_sales = sum(sales, na.rm = TRUE),
    total_profit = sum(profit, na.rm = TRUE),
    profit_margin = total_profit / total_sales
  ) %>%
  arrange(desc(profit_margin))

```

* Office Supplies like Labels, Paper, and Envelopes have the highest profit margins (> 40%).
* Some categories, such as Tables and Bookcases, show negative profitability.
* Profitability varies widely across sub-categories. Low-cost office supplies deliver strong margins, while some furniture items operate at a loss. This could be due to pricing issues, high return rates, or shipping costs. Product-level strategy adjustments may be needed.

**Visualize Profit Margin by Sub-Category:**
```{r profit margin by sub-category plot}
superstore_df %>%
  group_by(sub_category) %>%
  summarize(
    total_sales = sum(sales, na.rm = TRUE),
    total_profit = sum(profit, na.rm = TRUE),
    profit_margin = total_profit / total_sales
  ) %>%
  ggplot(aes(x = reorder(sub_category, profit_margin), y = profit_margin, fill = profit_margin)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Profit Margin by Sub-Category", x = "Sub-Category", y = "Profit Margin") +
  scale_fill_gradient2(low = "red", high = "green", midpoint = 0) +
  theme_minimal()

```

**Identify Loss-Making Categories:**
```{r negative-profit sub-categories}
superstore_df %>%
  group_by(sub_category) %>%
  summarize(total_profit = sum(profit, na.rm = TRUE)) %>%
  filter(total_profit < 0) %>%
  arrange(total_profit)

```

* Tables, Bookcases, and Supplies sub-categories have negative overall profit.
* These are clear loss-makers. Their continued inclusion in the catalog may require strategic review, particularly regarding pricing, supplier costs, or delivery/logistics challenges.

**Profitability by Region:**
```{r regional profitability}
superstore_df %>%
  group_by(region) %>%
  summarize(
    total_sales = sum(sales, na.rm = TRUE),
    total_profit = sum(profit, na.rm = TRUE),
    profit_margin = total_profit / total_sales
  ) %>%
  arrange(desc(profit_margin))

```

* The West region shows the highest profit margin (14.9%), while Central has the lowest (7.9%).
* Regional differences are significant. The West outperforms in both sales and margin, indicating stronger market dynamics or customer base. Central's lower margin may reflect operational inefficiencies or competitive pricing pressures.

# Key Business Insights

**1- Sales Are Concentrated in High-Value Items**
A small number of expensive products (e.g., Canon Copiers, Cisco Systems) generate a disproportionately high share of revenue. This highlights the importance of prioritizing inventory management, pricing strategies, and targeted marketing efforts around these "high-ticket" items.

**2- Customer Segments Differ by Value Type**
While the Consumer segment brings the highest total sales, the Home Office segment has the highest average profit margin. This suggests segmentation-based strategies: scale-focused for consumers, margin-focused for home office users.

**3- Sub-Category Profitability Is Highly Uneven**
Office Supplies like Labels and Paper offer exceptional margins (above 40%), while Furniture items like Tables and Bookcases result in financial losses. These insights point to opportunities for product rationalization and profit optimization.

**4- Negative-Profit Sub-Categories Need Attention**
Tables and Bookcases consistently generate losses. Root causes (e.g., high shipping cost, overstock, discounting) should be investigated, and pricing or portfolio strategies revised.

**5- West Region Leads in Profitability**
The West region achieves both high sales and the highest profit margin (14.9%), suggesting a robust customer base or effective operations. Conversely, the Central region shows underperformance and requires further analysis.

## Limitations

* The dataset reflects a limited time period, and seasonal trends cannot be assessed reliably.
* Gender, age, and income levels are not provided, limiting the depth of customer behavior analysis.
* Product lifecycles, pricing changes, or promotions are not included—key factors that could affect profitability.
* The impact of returns, refunds, or discounts on profitability is not directly observable.

# Conclusion 

This analysis uncovered key areas of opportunity for the business to optimize profitability and streamline its product portfolio. High-value technology items and office supplies drive revenue and margins, whereas select furniture categories require attention. Regional and segment-level patterns suggest targeted strategies to unlock further value. With improved data granularity and longitudinal tracking, future analysis can better inform pricing, marketing, and inventory decisions.

# Future Work
Future analyses could:
 
* analyze loss-making products by customer segment, region, and season to understand drivers,
* incorporate customer-level purchase patterns over time to assess CLV by segment,
* use time-series models to forecast demand and identify trends in top-performing categories,
* explore how discounts or promotions affect profit margins in different categories.

